---
title: "AED R and Friends: Working with databases using dplyr"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, echo=FALSE, warning=FALSE, purl=FALSE, message=FALSE}
options(repos="http://cran.rstudio.com/")
pkgs <- c("dplyr","tidyr","knitr","randomForest","caret","ranger",
          "party")
x<-lapply(pkgs, library, character.only = TRUE)
opts_chunk$set(tidy=T)
opts_chunk$set(root.dir = "/data/projects/aed_r/meetings")
set.seed(1)
```

# Basics of running random forests 
There are multiple ways to run random forests, we will look at three different packages for this: `randomForest` (the classic), `party` (new and improved), and `ranger` (the souped up classic).

## Data
Let's play with the NLA 2012 data!

```{r}
chla <- "https://www.epa.gov/sites/production/files/2016-11/nla2012_chla_wide.csv"
wq <- "https://www.epa.gov/sites/production/files/2016-12/nla2012_waterchem_wide.csv"
sites <- "https://www.epa.gov/sites/production/files/2016-12/nla2012_wide_siteinfo_08232016.csv"
secchi <- "https://www.epa.gov/sites/production/files/2016-12/nla2012_secchi_08232016.csv"
nla12<-full_join(read.csv(sites,stringsAsFactors = FALSE),
                 read.csv(wq,stringsAsFactors = FALSE)) %>%
  full_join(read.csv(chla,stringsAsFactors = FALSE)) %>%
  full_join(read.csv(secchi,stringsAsFactors = FALSE)) %>%
  filter(VISIT_NO == 1) %>%
  select(SITE_ID, AREA_HA,LAT_DD83,LON_DD83,PERIM_KM,ELEVATION,TURB_RESULT,DOC_RESULT,
         NTL_RESULT,PTL_RESULT,CHLX_RESULT,SECCHI)
names(nla12) <- tolower(names(nla12)) #Becuase I hate caps
nla12_cc <- nla12[complete.cases(nla12[,-1]),] %>%
  mutate(ntl_result = log1p(ntl_result), ptl_result = log1p(ptl_result), 
         chlx_result =  log1p(chlx_result))
tbl_df(nla12_cc)
```

## randomForest

First, set our seed so that results come out the same:

```{r}
# Set Seed
set.seed(1)
```

```{r}
library(randomForest)
nla_rf <- randomForest(chlx_result ~ ., data = nla12_cc[,-1], ntree = 1000 , 
                       mtry = 5, importance = T)
system.time(randomForest(chlx_result ~ ., data = nla12_cc[,-1],ntree = 1000,
                         mtry = 5, importance = T))
nla_rf
plot(nla_rf)
varImpPlot(nla_rf)
caret::varImp(nla_rf)

```

## party
```{r}
library(party)
nla_party <- cforest(chlx_result ~ ., data = nla12_cc[,-1], 
                     control = cforest_unbiased(ntree=1000))
system.time(cforest(chlx_result ~ ., data = nla12_cc[,-1], 
                     control = cforest_unbiased(ntree=1000)))
caret::cforestStats(nla_party)

```

## ranger
```{r}
library(ranger)
nla_ranger <- ranger(chlx_result ~ ., data = nla12_cc[,-1], num.trees = 1000, 
                     mtry = 5, importance = "impurity")
system.time(ranger(chlx_result ~ ., data = nla12_cc[,-1], num.trees = 1000, mtry = 5, 
                   importance = "impurity"))
nla_ranger
ranger::importance(nla_ranger)
```

# Compare predictions from 2012 models on 2007 data
My understanding of random forests is that separate test and training datasets are not needed to validate the model.  The resampling that occurs takes care of this and is another argument in favor of other methods that don't have this resampling and "out of bag" error built in.  That being said, I can see situations where it would be nice to see how well a model predicts an additional, independent dataset.  We (finally!!!) can do this with the NLA 2012 and 2007 data.

We will need to build our test dataset and since the `predict()` function is looking for datasets with the same strucutre we will make sure that happens.

## Build test datasets from 2007

```{r}
wc <- "https://www.epa.gov/sites/production/files/2014-10/nla2007_chemical_conditionestimates_20091123.csv"
sites <- "https://www.epa.gov/sites/production/files/2014-01/nla2007_sampledlakeinformation_20091113.csv"
secchi <- "https://www.epa.gov/sites/production/files/2014-10/nla2007_secchi_20091008.csv"
nla07 <- full_join(read.csv(wc, stringsAsFactors = TRUE),
                   read.csv(sites, stringsAsFactors = TRUE) , 
                   by = c("SITE_ID", "VISIT_NO")) %>%
  full_join(read.csv(secchi, stringsAsFactors = TRUE), 
            by = c("SITE_ID", "VISIT_NO")) %>%
  filter(VISIT_NO == 1) %>%
  select(site_id = SITE_ID, area_ha = AREA_HA, lat_dd83 = LAT_DD.x, 
         lon_dd83 = LON_DD.x, perim_km = LAKEPERIM, elevation = ELEV_PT, 
         turb_result =  TURB, doc_result = DOC, ntl_result = NTL, 
         ptl_result = PTL, chlx_result = CHLA, secchi = SECMEAN)
nla07_cc <- nla07[complete.cases(nla07[,-1]),] %>%
  mutate(ntl_result = log1p(ntl_result), ptl_result = log1p(ptl_result), 
         chlx_result =  log1p(chlx_result))
tbl_df(nla07_cc)
```

## Predict and compare

```{r}
nla07_chla_predict <- predict(object = nla_rf, newdata = nla07_cc)
plot(nla07_chla_predict,nla07_cc$chlx_result)
```
